Dispersion distance and the matter distribution of the Universe in dispersion space

### 2015-06-04 writeup/dm_space_distortions.tex
one more merger junk removal

The clustering of matter on large scales has been
heralded as the next great probe of the Universe after the cosmic microwave
background (CMB). The large-scale structure in principle contains far more
information than the CMB because it can be studied in three dimensions.
Traditionally the redshift of spectral lines, caused by the Hubble flow, has
been employed to measure radial distance and provide the third dimension.
Such redshift
surveys have now aggregated the positions of millions of galaxies into three
dimensional density maps, exposing a rich structure of clusters, filaments,
walls and voids.

@@

Neither redshift nor dispersion are perfect proxies for radial distance.
Redshifts are systematically biased by the peculiar velocities of objects
relative to the Hubble flow. This leads to additional apparent clustering
in redshift
space which was first calculated by \citet{1987MNRAS.227....1K}
using linear theory.
Similarly, as we show here, distance estimates from dispersion will be biased by
inhomogeneities in the electron density \citep{2014ApJ...780L..33M},
again leading to additional apparent
clustering of sources in dispersion space.  In this letter we calculate
these dispersion-space distortions and consider the detectability of the signal by
upcoming surveys.

@@

Large-scale structure is usually studied through its two-point statistics, most
commonly the power spectrum, $P(k)$.
Unlike the redshift-space distortions, Equation~\ref{e:delta_s} does not have a
simple form in harmonic space. The equation's third term couples harmonic modes
and thus the two-point statistics cannot be phrased as a simple power spectrum.
We will instead use $C^{ss}_\ell(\chi,\chi')$, which is
the cross-correlation angular power spectrum of the dispersion-space over-density,
on shells at $\chi$ and $\chi'$:
\begin{align}
    &C^{ss}_\ell(\chi, \chi') =
    \int\ud\Omega\ud\Omega' Y_{\ell m}(\hat n) Y^*_{\ell m}(\hat n')
    \langle \delta_{fs}(\hat n \chi) \delta_{fs}(\hat n' \chi')
        \rangle.
\end{align}
The first two terms in Equation~\ref{e:delta_s} are stationary. For stationary,
isotropic, tracers $x$ and $y$ we have 
$\langle \delta_x(\vec k, \chi) \delta_y(\vec k', \chi) \rangle = (2\pi)^3
\delta^3(\vec k - \vec k') P_{xy}(k, \chi)$.  If for a moment we ignore
structure evolution, the angular cross-power spectrum of such tracers is
\begin{align}
    %\delta_{\ell\ell'}\delta_{mm'}C^{xy}_\ell(\chi, \chi')
    %&= \int\ud\Omega\ud\Omega'Y_{\ell m}(\hat n) Y_{\ell' m'}(\hat n')
    %    \langle \delta_x(\hat n \chi) \delta_y(\hat n' \chi') \rangle,
    %    \\
    C^{xy}_\ell(\chi,\chi')
    &= \frac{2}{\pi}
\int_0^\infty\ud k k^2 j_\ell(k\chi) j_{\ell}(k\chi')P_{xy}(k).
\end{align}
In reality the power spectrum evolves on the order of a Hubble time.
The angular cross correlations will be very small unless $\chi$ and $\chi'$ are
within a few correlation lengths of one another, roughly a hundred
megaparsecs.  The evolution of the power spectrum is negligible over these time
differences which leads to a straight-forward way to include the evolution:
\begin{align}
    % I've tuned the alignment of the domain (using &)
    % here to get the equation number and
    % domain on the same line. Not elegant.
    C^{xy}_\ell(\chi,\chi') \approx&
        \frac{2}{\pi}
        \int_0^\infty\ud k k^2
        j_\ell(k\chi) j_{\ell}(k\chi')
        P_{xy}(k,(\chi + \chi')/2),\nonumber \\
    &|\chi - \chi'| \ll 1/aH.
\end{align}

>>>>

The clustering of matter on large scales has been
heralded as the next great probe of the Universe after the cosmic microwave
background (CMB). The large-scale structure in principle contains far more
information than the CMB because it can be studied in three dimensions.
Traditionally the redshift of spectral lines of `standard atoms', caused by the Hubble flow, has
been employed to measure radial distance and provide the third dimension.
Such redshift
surveys have now aggregated the positions of millions of galaxies into three
dimensional density maps, exposing a rich structure of clusters, filaments,
walls and voids.

@@

Neither redshift nor dispersion are perfect proxies for radial distance.
Redshifts are systematically biased by the peculiar velocities of objects
relative to the Hubble flow. This leads to additional apparent clustering
in redshift
space which was first calculated by \citet{1987MNRAS.227....1K}
using linear theory.
Similarly, as we show here, distance estimates from dispersion will be biased by
inhomogeneities in the electron density \citep{2014ApJ...780L..33M},
again leading to additional apparent
clustering of sources in dispersion space.  In this \emph{Letter} we calculate
these dispersion-space distortions and consider the detectability of the signal by
upcoming surveys.

@@

Large-scale structure is usually studied through its two-point statistics, most
commonly the power spectrum, $P(k)$.
Unlike the redshift-space distortions, Equation~\ref{e:delta_s} does not have a
simple form in harmonic space. The equation's third term couples harmonic modes
and thus the two-point statistics cannot be phrased as a simple power spectrum.
We will instead use $C^{ss}_\ell(\chi,\chi')$, which is
the cross-correlation angular power spectrum of the dispersion-space over-density,
on shells at $\chi$ and $\chi'$:
\begin{align}
    &C^{ss}_\ell(\chi, \chi') =
    \int\ud\Omega\ud\Omega' Y_{\ell m}(\hat n) Y^*_{\ell m}(\hat n')
    \langle \delta_{fs}(\hat n \chi) \delta_{fs}(\hat n' \chi')
        \rangle.
\end{align}
The first two terms in Equation~\ref{e:delta_s} are stationary. For stationary,
isotropic, tracers $x$ and $y$ we have 
$\langle \delta_x(\vec k, \chi) \delta_y(\vec k', \chi) \rangle = (2\pi)^3
\delta^3(\vec k - \vec k') P_{xy}(k, \chi)$.  If for a moment we ignore
structure evolution, the angular cross-power spectrum of such tracers is
\begin{align}
    %\delta_{\ell\ell'}\delta_{mm'}C^{xy}_\ell(\chi, \chi')
    %&= \int\ud\Omega\ud\Omega'Y_{\ell m}(\hat n) Y_{\ell' m'}(\hat n')
    %    \langle \delta_x(\hat n \chi) \delta_y(\hat n' \chi') \rangle,
    %    \\
    C^{xy}_\ell(\chi,\chi')
    &= \frac{2}{\pi}
\int_0^\infty\ud k k^2 j_\ell(k\chi) j_{\ell}(k\chi')P_{xy}(k).
\end{align}
In reality the power spectrum evolves on the order of a Hubble time.
The angular cross correlations will be very small unless $\chi$ and $\chi'$ are
within a few correlation lengths of one another, roughly a hundred
megaparsecs.  differences which leads to a straight-forward way to include the evolution of
the perturbations (essentially indistinguishable from the full expression):
\begin{align}
    % I've tuned the alignment of the domain (using &)
    % here to get the equation number and
    % domain on the same line. Not elegant.
    C^{xy}_\ell(\chi,\chi') \approx&
        \frac{2}{\pi}
        \int_0^\infty\ud k k^2
        j_\ell(k\chi) j_{\ell}(k\chi')
        P_{xy}(k,(\chi + \chi')/2),\nonumber \\
    &|\chi - \chi'| \ll 1/aH.
\end{align}
